The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-07-13 Paper/jaccardbaetica.tex
add new changes correction

To do this, a population approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}. Stamps will be used to identify economic patterns by analysing their similarity. If workshops and provinces share stamps with similar traits, then we can identify connections. By contrast, if we do not detect similar stamps between workshops, then workshops worked independently.

@@

The chronology of the workshops is wide spanning from the first to the third centuries AD
%\xavi{qué quiere decir que la cronología es diversa?} 
\citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. 
This fact is shown by the archaeological evidence that displays a highly specialised production with a long activity with apparently few changes\xavi{deberías desarrollar esto un poco más; qué tipo de evidencia demuestra esto? pocos cambios en qué? los talleres? las tipologías anfóricas?} \citep{remesal_anforas_2004}.

A large majority of this Baetican amphorae production is composed by Dressel 20 types\citep{dressel_ricerche_1878,millet_anforas_1998}. Dressel 20 amphorae were shipped and distributed from \textit{Baetica} throughout western Europe\xavi{esto ya lo has mencionado; quizás podrías explicar en 2 líneas la capacidad y forma de la dressel 20}.

@@

Dressel 20 was the amphora most stamped during the Roman Empire \citep[18]{millet_anforas_1998}. The usage period is associated with a large amphora production for almost three centuries\xavi{esto está repetido no?}.

The large sample size of recovered Dressel 20 stamps is distributed across all the western provinces of the Roman Empire. For this reason there are several publications of this type of evidence discussing its origin, its meaning, its long period of use and its spatial distribution~\citep{remesal_sellar_2016} (Fig.\ref{amphora}).\xavi{si dices que es muy popular debería haber aquí más citas aparte de Remesal}.

@@

The presence of Dressel 20 stamps in military camps in \textit{Britannia} has been widely studied in Roman archaeology \citep{carreras_britannia_1998}.\xavi{si dices widely studied no puedes referencias solo 1 autor}.

This intense consumption of olive oil also indicates a possible provincial structure designed to organize the supply of olive oil to military camps, such as \textit{Germania} (Remesal Rodrıguez, 1986)\xavi{por qué such as Germania? Aún no has explicado qué hacen en Germania}. There are no written records explaining how this redistribution of essential goods was organised in \textit{Britannia}, but the archaeological evidence suggests that cities may have been the central nodes of this local trade network~\citep[45]{funari_economic_2005}.

The consumption of olive oil would experience a progressive slowdown from the third century A.D. onwards. This date matches a change in market strategy in the Empire. This a gradual decrease on \textit{Baetican} olive oil exports can be observed as a decrease in the amount of Dressel 20 found in excavated sites as this typology is replaced by Dressel 23 amphorae \citep{rodriguez1991aceite,millet_anforas_1998}.\xavi{pero entonces la dressel 23 de donde viene? Cuidao que estás diciendo que se acaban las dressel 20, pero esto implica que hay menos aceite de oliva o que viene de otras partes o que va en recipientes distintos a la dressel 20?}

@@

\textit{Germania} presents a similar introduction pattern of olive oil than the one discussed for \textit{Britannia}. There is no archaeological evidence for olive oil consumption before the arrival of the Roman army to the region while a majority of recovered Dressel 20 amphorae are located in the military sites that formed the German \textit{limes}\xavi{cita?}.

@@

The goal of this study is exploring the effect of the production patterns between different centres. We are especially interested in identifying links between production and consumption centres by using amphoric stamps. To do that, we use the CEIPAC database to collect stamps from different places. The CEIPAC dataset contains over 50.000 of epigraphy records found in amphorae, mostly from \textit{Monte Testaccio}.
This study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. A way to analyse is to use a quantitative framework to measure the similarity between amphoric stamps. Here we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potential spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance. 
\xavi{estos 2 párrafos explican objetivos, que deberían ir en la intro y no aquí, donde deberías empezar a hablar directamente de la BBDD}

@@

To study the stamps found in the Baetican workshops we used the data collected in the CEIPAC database of amphoric epigraphy~\citep{remesal_centro_2015} (see the database here \url{http://romanopendata.eu}). This database contains over 50.000 epigraphy records found on different types of amphorae\xavi{podrías mover esta parte a los 2 párrafos anteriores y explicar la bbdd brevemente antes de empezar a hablar de la Bética (porque es info que también aplica a los otros 2 casos de estudio no?}.

@@

Our dataset also contained a new categorical variable defining the \textit{conventus} of each site. The \textit{conventus} were administrative centres for territorial organisation in the Roman Empire \citep[58]{ozcariz_gil_administracion_2013}\xavi{estaría bien ampliar un poco esto; qué organizaban exactamente?}. The production area for Dressel 20 amphorae extended across three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter Hispalis), \textit{Cordubensis} (currently C\'ordoba, hereafter Corduba) and \textit{Astigi} (currently Écija, Sevilla, hereafter Astigi) \citep{rodriguez_economioleicola_1977,chicdatos2001,berni_millet_epigrafianforica_2008} .

@@

It is important to mention that some workshops exhibited the same coordinates but they were catalogued as different workshops in Roman Studies (i.e. Tesorillo de Doña Mencia and Doña Mencia). The analysis took into account this particularity but some biases can occur due to the archaeological dataset\xavi{No se entiende; deberías explicarlo un poco mejor y citar algo si dices eso. Piensa que la mayoría de lectores no sabrán quién es Doña Mencia}.

@@

The distribution of amphora stamps for each \textit{conventus} can be seen in Figure~\ref{frequency}. The majority of stamps are concentrated in \textit{Hispalis} (574 stamps) while \textit{Corduba} and \textit{Astigi} have roughly half this sample size (267 and 146 stamps). The workshops of the three conventus show a similar pattern on the stamp frequency distribution with the exceptions of 2 large workshops in \textit{Hispalis} (La Catria and Arva). On these two workshops a comparatively large amount of stamps were found (29 different code stamps on each of them)\xavi{una cosa, aquí has pasado de sellos a códigos, y los valores son distintos que en el gráfico anterior. Deberías clarificarlo antes de empezar a hablar de la figura}. According to previous studies those workshops became the most important centres of amphora production in the region\xavi{cita de estos estudios}. It is worth mentioning that a majority of these stamps were collected during ield surveys with no excavation, so this difference in sample size could be biased due to differences in intensity across the different sites~\citep{arva_1997}.

@@

Considering our dataset as a non-uniform sample, this method provides a useful tool to handle large samples with different sizes and diversity \citep{wolda_similarity_1981}. Morisita-Horn index gives a value between 0 (sites have an exact stamp codes and frequency distribution) and 1 (complete difference between stamp codes). To apply Morisita-Horn we firstly calculated the number of times that each stamp code appeared in an amphora workshop. This method allowed us to bear in mind a similar number of times for each repeated stamp per workshop\xavi{no entiendo esta frase}. If two workshops had similar stamp code distribution then the distance index would be close to 0 whereas sites with completely unrelated stamp codes would give a distance close to 1.

@@

This work has explored to what extent the distribution of amphoric stamps can provide insights on the organisation of olive oil Roman markets both at the production and consumption areas. For this reason, an index of dissimilarity was used to detect differences between the distribution of the amphoric stamps and the spatial distance of producers and consumption centres. The general purpose was to explore if such differences found in the stamp codes could play an important role in the Roman market\xavi{este párrafo es un poco reiterativo no?}.

@@

@@

Both consumption areas showed a correlation between spatial distance of sites and their similarity of amphoric stamps. In the case of  \textit{Britannia}, the correlation was higher than \textit{Germania}. In \textit{Britannia} the majority of similarity stamps were mainly found in military centres. This could be also interpreted by an intensity bias where military centres have been mostly excavated than civil centres\xavi{pero no decías que el aceite llegaba por las legione? Entonces por qué es un bias? En todo caso ponlo al final de la sección y no al inicio}.

@@

.The areas mostly militarised share similar stamps than civil areas. However, we do not detect a concrete pattern regarding the distribution of the stamps in the German limes \citep{xanten2018}.

It is worth mentioning that it was not detected a robust model of organisation in both cases with regard to the distribution of stamps in consumption centres\xavi{de nuevo esto es redundante no?}. This evidence means that it is unknown whether some production centres went to one province or another or, at least, that they can be clearly reflected in the data with a greater similarity in the amphoric stamps. Thus, production centres could have distributed randomly olive oil both \textit{Britannia} and \textit{Germania}. Neither we do not detect production centres dedicated to the distribution of olive oil in a certain province.

@@

Alternatively, the presence of different stamps in the same workshop could imply some kind of organisation from within the workshop only affecting nearby centres \citep{juanmorostesis}. The stamped amphorae could have allowed potters to organise the different production batches for posterior commercialisation\xavi{estos 2 párrafos igual deberían ir a la discusión no? La conclusión debería ser algo más general y aquí estás aún interpretando resultados}.

>>>>

This study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. To do this, a population approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}.  A way to analyse is to use a quantitative framework to measure the similarity between amphoric stamps. Here we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potential spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance. Stamps will be used to identify economic patterns by analysing their similarity. If workshops and provinces share stamps with similar traits, then we can identify connections. By contrast, if we do not detect similar stamps between workshops, then workshops worked independently.

@@

The chronology of the workshops is wide spanning from the first to the third centuries AD
%\xavi{qué quiere decir que la cronología es diversa?} 
\citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. 
This fact is shown by the archaeological evidence that displays a highly specialised production with a long activity \citep{remesal_anforas_2004}. Thus, the specialisation of this production made amphorae with apparently few changes in their forms, reaching a high degree of standardisation.

%\xavi{deberías desarrollar esto un poco más; qué tipo de evidencia demuestra esto? pocos cambios en qué? los talleres? las tipologías anfóricas?} .

A large majority of this Baetican amphorae production is composed by Dressel 20 types\citep{dressel_ricerche_1878,millet_anforas_1998}. Dressel 20 amphorae have a recognisable globular form with short neck and oval-shaped handles. Its  capacity allowed to hold 70-75 litres of olive oil to be shipped and distributed throughout Empire \citep{berni_dressel_2016}.

%were shipped and distributed from \textit{Baetica} throughout western Europe. 
%\xavi{esto ya lo has mencionado; quizás podrías explicar en 2 líneas la capacidad y forma de la dressel 20}.

@@

Dressel 20 was the amphora most stamped during the Roman Empire \citep[18]{millet_anforas_1998}.

%The usage period is associated with a large amphora production for almost three centuries\xavi{esto está repetido no?}.

The large sample size of recovered Dressel 20 stamps is distributed across all the western provinces of the Roman Empire. For this reason there are several publications of this type of evidence discussing its origin, its meaning, its long period of use and its spatial distribution~\citep{dressel_ricerche_1878,
rodriguez_economioleicola_1977,
chicepi1985,millet_anforas_1998, remesal_sellar_2016} (Fig.\ref{amphora}).

%\xavi{si dices que es muy popular debería haber aquí más citas aparte de Remesal}.

@@

The presence of Dressel 20 stamps in military camps in \textit{Britannia} has been widely studied in Roman archaeology \citep{williams_importation_1983,
funari_corpus_1996,
carreras_britannia_1998,carreras_abastecimiento_2003}.%\xavi{si dices widely studied no puedes referencias solo 1 autor}.

This intense consumption of olive oil also indicates a possible provincial structure designed to organize the supply of olive oil to military camps,

%such as \textit{Germania} (Remesal Rodrıguez, 1986)\xavi{por qué such as Germania? Aún no has explicado qué hacen en Germania}.

There are no written records explaining how this redistribution of essential goods was organised in \textit{Britannia}, but the archaeological evidence suggests that cities may have been the central nodes of this local trade network~\citep[45]{funari_economic_2005}.

The consumption of olive oil would experience a progressive slowdown from the third century A.D. onwards. This date matches a change in market strategy in the Empire. This a gradual decrease on \textit{Baetican} olive oil exports can be observed as a decrease in the amount of Dressel 20 found in excavated sites as this typology is replaced by other types of amphorae from other places \citep{rodriguez1991aceite,millet_anforas_1998}.

%\xavi{pero entonces la dressel 23 de donde viene? Cuidao que estás diciendo que se acaban las dressel 20, pero esto implica que hay menos aceite de oliva o que viene de otras partes o que va en recipientes distintos a la dressel 20?}

@@

\textit{Germania} presents a similar introduction pattern of olive oil than the one discussed for \textit{Britannia}. There is no archaeological evidence for olive oil consumption before the arrival of the Roman army to the region while a majority of recovered Dressel 20 amphorae are located in the military sites that formed the German \textit{limes} \citep{remesal_germaniaengl_2002}.

%\xavi{cita?}

@@

The goal of this study is exploring the effect of the production patterns between different centres. We are especially interested in identifying links between production and consumption centres by using amphoric stamps.

%To do that, we use the CEIPAC database to collect stamps from different places. The CEIPAC dataset contains over 50.000 of epigraphy records found in amphorae, mostly from \textit{Monte Testaccio}.

To study the amphoric stamps, we used the data collected in the CEIPAC database of amphoric epigraphy from different places~\citep{remesal_centro_2015} (see the database here \url{http://romanopendata.eu}). The CEIPAC dataset contains over 50.000 epigraphy records found on different types of amphorae, mostly from \textit{Monte Testaccio}.

%\xavi{podrías mover esta parte a los 2 párrafos anteriores y explicar la bbdd brevemente antes de empezar a hablar de la Bética (porque es info que también aplica a los otros 2 casos de estudio no?}.

%This study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. A way to analyse is to use a quantitative framework to measure the similarity between amphoric stamps. Here we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potential spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance. 
%\xavi{estos 2 párrafos explican objetivos, que deberían ir en la intro y no aquí, donde deberías empezar a hablar directamente de la BBDD}

@@

@@

Our dataset also contained a new categorical variable defining the \textit{conventus} of each site. The \textit{conventus} were administrative centres for territorial organisation under the Roman Empire \citep[58]{ozcariz_gil_administracion_2013}.
The creation of the \textit{conventus} in some Roman provinces was a form of control with the purpose of organising the administration in the provinces \citep{albertini_les_1923}.

%\xavi{estaría bien ampliar un poco esto; qué organizaban exactamente?}

The production area for Dressel 20 amphorae extended across three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter Hispalis), \textit{Cordubensis} (currently C\'ordoba, hereafter Corduba) and \textit{Astigi} (currently Écija, Sevilla, hereafter Astigi) \citep{rodriguez_economioleicola_1977,chicdatos2001,berni_millet_epigrafianforica_2008} .

@@

It is important to mention that some workshops exhibited the same coordinates but they were catalogued as different workshops in Roman Studies (i.e. nearby amphora workshops in \textit{Baetica} named as Tesorillo de Doña Mencia and Doña Mencia). The analysis took into account this particularity but some biases can occur due to the archaeological dataset.

\xavi{No se entiende; deberías explicarlo un poco mejor y citar algo si dices eso. Piensa que la mayoría de lectores no sabrán quién es Doña Mencia}.

@@

The distribution of code stamps for each \textit{conventus} can be seen in Figure~\ref{frequency}. The majority of stamps are concentrated in \textit{Hispalis} (574 code stamps) while \textit{Corduba} and \textit{Astigi} have roughly half this sample size (267 and 146 code stamps). The workshops of the three \textit{conventus} show a similar pattern on the stamp frequency distribution with the exceptions of 2 large workshops in \textit{Hispalis} (La Catria and Arva). On these two workshops a comparatively large amount of code stamps were found (29 different code stamps on each of them)

%\xavi{una cosa, aquí has pasado de sellos a códigos, y los valores son distintos que en el gráfico anterior. Deberías clarificarlo antes de empezar a hablar de la figura}.

According to previous studies those workshops became the most important centres of amphora production in the region\xavi{cita de estos estudios}. It is worth mentioning that a majority of these stamps were collected during ield surveys with no excavation, so this difference in sample size could be biased due to differences in intensity across the different sites~\citep{arva_1997}.

@@

Considering our dataset as a non-uniform sample, this method provides a useful tool to handle large samples with different sizes and diversity \citep{wolda_similarity_1981}. Morisita-Horn index gives a value between 0 (sites have an exact stamp codes and frequency distribution) and 1 (complete difference between stamp codes). To apply Morisita-Horn we firstly calculated the number of times that each stamp code appeared in an amphora workshop. 
This method allowed us to bear in mind the number of frequency of stamps for each workshop. 
%This method allowed us to bear in mind a similar number of times for each repeated stamp per workshop\xavi{no entiendo esta frase}.

If two workshops had similar stamp code distribution then the distance index would be close to 0 whereas sites with completely unrelated stamp codes would give a distance close to 1.

@@

This work has explored to what extent the distribution of amphoric stamps can provide insights on the organisation of olive oil Roman markets both at the production and consumption areas. For this reason, an index of dissimilarity was used to detect differences between the distribution of the amphoric stamps and the spatial distance of producers and consumption centres.

%The general purpose was to explore if such differences found in the stamp codes could play an important role in the Roman market\xavi{este párrafo es un poco reiterativo no?}.

@@

Alternatively, the presence of different stamps in the same workshop could imply some kind of organisation from within the workshop only affecting nearby centres \citep{juanmorostesis}. The stamped amphorae could have allowed potters to organise the different production batches for posterior commercialisation.
%\xavi{estos 2 párrafos igual deberían ir a la discusión no? La conclusión debería ser algo más general y aquí estás aún interpretando resultados}.

@@

Both consumption areas showed a correlation between spatial distance of sites and their similarity of amphoric stamps. In the case of  \textit{Britannia}, the correlation was higher than \textit{Germania}. In \textit{Britannia} the majority of similarity stamps were mainly found in military centres.

@@

The areas mostly militarised share similar stamps than civil areas. However, we do not detect a concrete pattern regarding the distribution of the stamps in the German limes \citep{xanten2018}. This could be also interpreted by an intensity bias where military centres have been mostly excavated than civil centres

%\xavi{pero no decías que el aceite llegaba por las legione? Entonces por qué es un bias? En todo caso ponlo al final de la sección y no al inicio}.

%It is worth mentioning that it was not detected a robust model of organisation in both cases with regard to the distribution of stamps in consumption centres\xavi{de nuevo esto es redundante no?}.

This evidence means that it is unknown whether some production centres went to one province or another or, at least, that they can be clearly reflected in the data with a greater similarity in the amphoric stamps. Thus, production centres could have distributed randomly olive oil both \textit{Britannia} and \textit{Germania}. Neither we do not detect production centres dedicated to the distribution of olive oil in a certain province.

@@

%Alternatively, the presence of different stamps in the same workshop could imply some kind of organisation from within the workshop only affecting nearby centres \citep{juanmorostesis}. The stamped amphorae could have allowed potters to organise the different production batches for posterior commercialisation\xavi{estos 2 párrafos igual deberían ir a la discusión no? La conclusión debería ser algo más general y aquí estás aún interpretando resultados}.
